### 2017-07-04 reasoning.tex
Weak isolation example added.

We now describe a proof system that lets us prove the correctness of a \txnimp
program $c$ w.r.t its high-level invariant $I$, on a machine that
satisfies the isolation specifications ($\I$) of its
transactions\footnote{Note the difference between $I$ and $\I$. The
former constitute \emph{proof} \emph{obligations} for the programmer,
whereas the latter describes a transaction's \emph{assumptions} about
operational characteristics of the underlying machine.}.  Our proof
system is essentially an adapatation of the rely-guarantee reasoning
framework to the setting of weakly isolated database transactions. It
deals with the additional complexity introduced by this setting, while
also taking advantage of the opportunities it provides to simplify the
reasoning. The primary challenge is the weak isolation; how do we
relate a transaction's isolation specification ($\I$) to its rely
relation ($R$) that describes the environment, so that the
interference is considered only insofar as the isolation level allows
it? Our proof system shows how to compute a rely relation ($R$) modulo
the isolation specification ($\I$) of a transaction. Another
characteristic of the transaction setting is the atomicity of a
trasaction's aggregate writes; a transaction's writes are not visible
to its peers until it commits. In the context of rely-guarantee, this
means that the transactions guarantee ($G$) should capture the
aggregate effect of a transaction, and not its individual writes.
While \C{atomic} blocks are also present in the shared-memory
programs, the fact that a transactions are weakly isolated introduces
complexity.  Unlike an \C{atomic} block, the effect of a transaction
is \emph{not} a sequential composition of the effects of its
statements because each statement witnesses a potentially different
version of the state. This is illustrated in
Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
\C{x} can be composed to describe the effect of an \C{atomic} block,
while same cannot be done in a transaction. Thus, weak isolation needs
to be taken into account when defining and verifying guarantees, and
our proof system shows how.

>>>>

\begin{figure}
\centering
\begin{tabular}{l|l}
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  atomic {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{X = k! \conj Y = X\}
\end{decoration}$
\end{txnimpcode}
&
\begin{txnimpcode}
$\begin{decoration}
  \{X = k\}
\end{decoration}$
  transaction {
    v1 := 0; v2 := 1;
    while(v1<X) {
      v2 := v2 * (X-v1);
      v2 := v2 * v1;
      v1 := v1+1;
    }
    X := v2; Y := v2;
  }
$\begin{decoration}
  \{Y = X\}
\end{decoration}$
\end{txnimpcode}
\end{tabular}
  \caption{Factorial program written as an atomic block and a weakly
  isolated transaction (shared variables have uppercase names).
  Effect of multiple iteratons of the \C{atomic} loop can be
  summarized as $k!$, while same cannot be done with the
  \C{transaction} loop, where weak isolation allows reads in different
  iterations to return different values for $X$.  Nonetheless,
  post-state satisfies $X=Y$ in both cases due to atomicity of
  writes.}
\label{fig:atomic-vs-transaction}
\end{figure}

We now describe a proof system that lets us prove the correctness of a \txnimp
program $c$ w.r.t its high-level invariant $I$, on a machine that
satisfies the isolation specifications ($\I$) of its
transactions\footnote{Note the difference between $I$ and $\I$. The
former constitute \emph{proof} \emph{obligations} for the programmer,
whereas the latter describes a transaction's \emph{assumptions} about
operational characteristics of the underlying machine.}.  Our proof
system is essentially an adapatation of the rely-guarantee reasoning
framework to the setting of weakly isolated database transactions. It
deals with the additional complexity introduced by this setting, while
also taking advantage of the opportunities it provides to simplify the
reasoning. The primary challenge is the weak isolation; how do we
relate a transaction's isolation specification ($\I$) to its rely
relation ($R$) that describes the environment, so that the
interference is considered only insofar as the isolation level allows
it? Our proof system shows how to compute a rely relation ($R$) modulo
the isolation specification ($\I$) of a transaction. Another
characteristic of the transaction setting is the atomicity of a
trasaction's aggregate writes; a transaction's writes are not visible
to its peers until it commits. In the context of rely-guarantee, this
means that the transactions guarantee ($G$) should capture the
aggregate effect of a transaction, and not its individual writes.
While \C{atomic} blocks are also present in the shared-memory
programs, the fact that a transactions are weakly isolated introduces
complexity.  Unlike an \C{atomic} block, the effect of a transaction
is \emph{not} a sequential composition of the effects of its
statements because each statement witnesses a potentially different
version of the state. This is illustrated in
Fig.~\ref{fig:atomic-vs-transaction}, where a series of operations on
\C{X} can be composed to describe the effect of an \C{atomic} block,
while same cannot be done in a transaction. Thus, weak isolation needs
to be taken into account when defining and verifying guarantees, and
our proof system shows how.
